---
title: "Cross-Species HVG-CCA: A Step-by-Step Tutorial"
author: "chilijuido23"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
vignette: >
  %\VignetteIndexEntry{Cross-Species HVG-CCA Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

# Overview

This tutorial walks through the complete pipeline for integrating single-cell
RNA-seq data across species using **adaptive HVG-based ortholog alignment**.

By the end, you will:
- Understand why standard ortholog alignment fails for fish species
- Preprocess scRNA-seq data from multiple species
- Run adaptive gene alignment
- Perform Seurat CCA integration
- Generate publication-quality visualizations

# The Problem We're Solving

## Why Can't We Just Use Ortholog Tables?

Consider this scenario: you have scRNA-seq data from human, mouse, lizard,
zebrafish, and large yellow croaker hypothalamus. You want to integrate them
to find conserved cell types.

Standard approach: use Ensembl BioMart to download orthologous gene pairs.

**The problem**: Zebrafish has `sox2a` AND `sox2b`. Human has `SOX2`.
Which zebrafish gene should map to `SOX2`?

```{r echo=FALSE, fig.cap="Multi-copy gene dilemma"}
# Illustration
library(ggplot2)
df <- data.frame(
  species = rep(c("Human", "Mouse", "Lizard", "Zebrafish", "Croaker"), each = 1),
  gene = c("SOX2", "Sox2", "SOX2", "sox2a\nsox2b", "sox2a\nsox2b"),
  y = c(1, 1, 1, 1, 1)
)
ggplot(df, aes(x = species, y = y, label = gene)) +
  geom_point(size = 5, color = "steelblue") +
  geom_text(vjust = -1, size = 4) +
  ylim(0.5, 1.8) +
  labs(title = "Gene Name Ambiguity Across Species",
       subtitle = "Fish have multiple paralogs for each mammalian gene") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Our solution: **let expression data decide which paralog to use**.

# Step 1: Preprocessing

```{r eval=FALSE}
# Run in terminal:
cd scripts
Rscript 01_preprocessing.R \
  --species_config species_config.yaml \
  --outdir ../data \
  --min_features 200 \
  --max_mt_percent 25
```

Inside R, the preprocessing looks like this:

```{r eval=FALSE}
library(Seurat)

# Load 10x data for each species
human_10x <- Read10X("path/to/human/filtered_feature_bc_matrix")
human <- CreateSeuratObject(human_10x, project = "human")

# QC and filter
human[["percent.mt"]] <- PercentageFeatureSet(human, pattern = "^MT-")
human <- subset(human, subset = nFeature_RNA > 200 & percent.mt < 25)

# Normalize
human <- NormalizeData(human)

# Repeat for all species...
```

# Step 2: Adaptive Ortholog Alignment (The Core Innovation)

This is where the magic happens:

```{r eval=FALSE}
# Terminal:
Rscript 02_adaptive_ortholog_alignment.R \
  --input ../data/all_species_preprocessed.rds \
  --n_hvg 2000 \
  --outdir ../data
```

What happens under the hood:

```{r eval=FALSE}
source("../R/adaptive_ortholog.R")

# 1. Each species independently computes top 2000 HVGs
# 2. Single-copy species → UPPERCASE
# 3. Multi-copy species → for each gene family, pick the highest-expressed paralog
# 4. Intersect all gene sets

align_result <- adaptive_ortholog_align(
  seurat_objects = seurat_list,
  species_info = list(
    human    = list(copy_type = "single"),
    mouse    = list(copy_type = "single"),
    lizard   = list(copy_type = "single"),
    zebrafish = list(copy_type = "multi",
                     copy_pattern = "^(\w+?)([a-z]\d*|[a-z]?)$"),
    croaker  = list(copy_type = "multi",
                    copy_pattern = "^(\w+?)([a-z]\d*|[a-z]?)$")
  ),
  n_hvg = 2000
)

# See which paralogs were selected
head(align_result$multi_copy_choices$zebrafish)
```

### Understanding the Paralog Selection

For zebrafish, each HVG that belongs to a multi-copy family is ranked by
mean expression across all cells:

| Gene Family | Paralogs        | Mean Expression | Selected |
|-------------|-----------------|-----------------|----------|
| sox2        | sox2a, sox2b   | 3.2 vs 0.8     | sox2a    |
| neurod1     | neurod1a, neurod1b | 2.1 vs 1.9 | neurod1a |
| gata3       | gata3a, gata3b | 1.5 vs 4.7     | gata3b   |

The logic: if `sox2a` is expressed 4× higher than `sox2b` in these cells,
it's likely the functional equivalent of mammalian SOX2 *in this tissue*.

# Step 3: CCA Integration

```{r eval=FALSE}
Rscript 03_cca_integration.R \
  --input ../data/all_species_preprocessed.rds \
  --align_result ../data/align_result.rds \
  --n_dim 30
```

This step:
1. Renames genes in each species to the unified aligned gene set
2. Merges all species into one Seurat object
3. Runs Seurat's CCA-based integration pipeline
4. Computes UMAP and clustering

# Step 4: Visualization

```{r eval=FALSE}
Rscript 04_visualization.R \
  --input ../data/integrated_seurat.rds \
  --outdir ../figures
```

# Interpreting Results

```{r eval=FALSE}
integrated <- readRDS("../data/integrated_seurat.rds")

# Check integration quality
DimPlot(integrated, group.by = "species")

# Are species mixing well? Good integration shows:
# - Cells from different species intermingled (not separated by species)
# - Clusters contain cells from multiple species
# - CCA shows smooth transitions, not isolated species islands

# Species composition per cluster
table(integrated$seurat_clusters, integrated$species)

# Find conserved markers
markers <- FindConservedMarkers(
  integrated,
  ident.1 = 0,  # cluster of interest
  grouping.var = "species"
)
```

# Troubleshooting

## "Very few aligned genes (<50)"

Try:
1. Reduce `n_hvg` to 1500 — fewer genes per species = larger intersection
2. Check your `copy_pattern` regex — are paralogs being correctly grouped?
3. Verify gene name conventions — are all single-copy species truly uppercase?

## "Species don't mix on UMAP"

Try:
1. Increase `n_dim` in CCA (30 → 50)
2. Verify that the aligned gene set is truly informative (check HVG overlap)
3. Consider adding `k.weight` parameter in `FindIntegrationAnchors()`

## "Integration is slow / runs out of memory"

Try:
1. Downsample each species to ~5000 cells before integration
2. Reduce `n_hvg` to 1000
3. Use `k.filter` parameter to limit anchors per cell

# Session Info

```{r}
sessionInfo()
```
